---
title: "Explainer: Measuring Energy in Population-Based Metaheuristics"
author: "JJ Merelo & Cecilia Merelo-Molina"
date: "`r Sys.Date()`"
output: bookdown::html_document2
bibliography: [ours.bib, energy.bib, ga-energy.bib, GAs.bib, julia.bib, metaheuristics.bib]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
library(dplyr)
library(kableExtra)

source("energyR/R/process_deltas.R") # Our own utility library
source("energyR/R/process_normal_deltas.R") # Our own utility library

create_summary <- function(data) {
  return(
    data %>%
      group_by(dimension, population_size, max_gens) %>%
      summarise(
        mean_delta_PKG = mean(delta_PKG),
        median_delta_PKG = median(delta_PKG),
        trimmed_mean_delta_PKG = mean(delta_PKG, trim=0.2),
        sd_delta_PKG = sd(delta_PKG),
        iqr_delta_PKG = IQR(delta_PKG),
        iqr_PKG = IQR(PKG),
        conf_interval_delta_PKG = sprintf("[%s, %s]", 
                                          round(t.test(delta_PKG)$conf.int[1], 2), 
                                          round(t.test(delta_PKG)$conf.int[2], 2)),
        .groups = 'drop'
      )
  )
}
```

This research focuses on the Brave New Algorithm (BNA) applied to the Sphere function to establish a methodology for profiling energy consumption. We use the Intel RAPL interface to measure the package-level energy (PKG) in Joules, seeking to isolate the actual workload from system overhead.

In this context, the **baseline** refers to the energy consumed by the system while idle or running only background processes. The **delta** PKG represents the estimated energy consumption of the algorithm itself, calculated by subtracting the baseline from the total measured energy. The parameter **max_gens** (maximum number of generations without improvement) serves as the stopping criterion for the evolutionary process.

# Introduction: Can we measure energy consumption in population-based metaheuristics?

Energy efficiency is becoming a critical engineering requirement for metaheuristics, yet measuring it accurately on modern multi-tasking systems is challenging. Modern processors employ dynamic power management that introduces high variability into measurements. This explainer details a methodology to extract reliable energy profiles using experimental design and statistical processing.

# Why a simple baseline is not enough

Does background energy consumption remain constant enough for a single baseline measurement?

```{r baseline-variability, fig.cap="PKG energy measured over time for two separate baseline measurement sets."}
baseline_ola_100s <- read.csv("data/ola-base-ola-baseline-14-Dec-12-06-42.csv")
baseline_ola_100s$cum_seconds <- cumsum(baseline_ola_100s$seconds)
baseline_ola_100s$group <- "First experiment"

baseline_ola_v2_data <- read.csv("data/ola-1.11.7-v2-baseline-v2-14-Dec-20-40-47.csv")
baseline_ola_v2_data$cum_seconds <- cumsum(baseline_ola_v2_data$seconds)
baseline_ola_v2_data$group <- "Second experiment"

baseline_both <- rbind(baseline_ola_100s, baseline_ola_v2_data)

ggplot(baseline_both, aes(x = cum_seconds, y = PKG, group=group, color=group)) +
    geom_line() +
    labs(x = "Cumulative Time (s)", y = "Energy Consumption (Joules)") +
    theme_minimal()
```

The baseline energy levels shift significantly over time and between different experimental sessions, sometimes by as much as 100 Joules. Even within a single session, background noise is not uniform, which can lead to "negative" energy estimates if a static baseline is subtracted from a workload run during a low-activity period.

**Takeaway:** Static baselines are insufficient because system state is non-stationary; measurements must account for temporal drift.

# The issue with separated baseline and workload experiments

How can we mitigate temporal drift in energy measurements?

```{r workload-segments-original, fig.cap="PKG energy measured vs. time for workload measurements. Dashed lines indicate baseline averages; colors distinguish configurations."}
workload_data_ola <- read.csv("data/ola-1.11.7-ola-14-Dec-13-02-30.csv")
# Use the same baseline summary as in original context
baseline_summary <- read.csv("data/ola-base-ola-baseline-14-Dec-12-06-42.csv") %>%
      group_by(dimension, population_size) %>%
      summarise(trimmed_mean_PKG = mean(PKG, trim=0.2), .groups = 'drop')

workload_data_ola$cum_seconds <- cumsum(workload_data_ola$seconds)
workload_data_ola$config <- paste0(workload_data_ola$dimension, "D, P=", workload_data_ola$population_size)

# Create color palette
n_configs <- length(unique(workload_data_ola$config))
colors <- scales::hue_pal()(n_configs)
names(colors) <- unique(workload_data_ola$config)

plot_with_segments <- ggplot(workload_data_ola, aes(x = cum_seconds, y = PKG, color = config)) +
    geom_line(alpha=0.6) +
    labs(x = "Cumulative Time (s)", y = "Energy Consumption (Joules)", color="Config") +
    theme_minimal() +
    ylim(300, 750) +
    scale_color_manual(values=colors)

# Add dashed lines for baseline averages
for (i in 1:nrow(baseline_summary)) {
  dim <- baseline_summary$dimension[i]
  pop <- baseline_summary$population_size[i]
  config_name <- paste0(dim, "D, P=", pop)
  
  # Filter data for this segment
  seg_data <- workload_data_ola[workload_data_ola$dimension == dim & workload_data_ola$population_size == pop, ]
  
  if (nrow(seg_data) > 0) {
    plot_with_segments <- plot_with_segments +
      geom_segment(x = min(seg_data$cum_seconds), y = baseline_summary$trimmed_mean_PKG[i],
                   xend = max(seg_data$cum_seconds), yend = baseline_summary$trimmed_mean_PKG[i], 
                   linetype="dashed", color=colors[config_name], linewidth=1)
  }
}
print(plot_with_segments)
```

But this effectively leads to a confusion of the energy measured for different parametrizations, which are almost impossible to distinguish.

```{r workload-delta, fig.cap="Figure 3: Workload measurements (processed delta energy)."}
# Process the data to get the delta PKG
workload_data_delta <- process_normal_deltas(read.csv("data/ola-1.11.7-ola-14-Dec-13-02-30.csv"))
ggplot(workload_data_delta, 
       aes(x = delta_seconds, y = delta_PKG, color=interaction(dimension,population_size))) +
      geom_point() +
      labs(x = "Delta Time (s)", y = "Delta Energy Consumption (Joules)") +
      theme_minimal()
```

# The Sequential Mixed Strategy

The sequential mixed strategy proposes workload and baseline in a single run, making the state in which the two processes are run close enough to be effectively the same (most of the times, at least)

```{r sequential-mixed, fig.cap="Sequential baseline and workload measurements showing tracking of system state."}
ola_mixed_no0 <- read.csv("data/ola-1.11.8-ola-no0-16-Dec-17-43-49.csv")
ola_mixed_no0$cumulative_time <- cumsum(ola_mixed_no0$seconds)

ggplot(ola_mixed_no0, aes(x = cumulative_time, y = PKG, group=work, color=work)) +
    geom_line() +
    labs(x = "Cumulative Time (s)", y = "Energy Consumption (Joules)") +
    theme_minimal()

ola_no0_base_PKG <- ola_mixed_no0[ ola_mixed_no0$work == "base-ola-no0",]$PKG
ola_no0_workload_PKG <- ola_mixed_no0[ ola_mixed_no0$work == "ola-no0",]$PKG
initial_correlation = cor(ola_no0_base_PKG, ola_no0_workload_PKG, use = "complete.obs")
```

The correlation between the energy yielded by baseline and workload runs is now `r initial_correlation`, which is a moderate correlation. In general, baseline and workload will change in the same way, implying that they are tracking system state. 

**Takeaway:** By interleaving baseline and workload runs in a sequential manner, we ensure that both measurements capture similar system states, effectively cancelling out long-term temporal drift.

Our initial assumption is that this would dampen noise. Let's see

```{r workload, fig.cap="Workload distribution in this case"}
mixed_workload_delta <- process_normal_deltas(read.csv("data/ola-1.11.7-mixed-ola-mixed-15-Dec-19-49-11.csv"))
ggplot(mixed_workload_delta, 
       aes(x = delta_seconds, y = delta_PKG, color=interaction(dimension,population_size))) +
      geom_point( alpha=0.5) +
      labs(x = "Delta Time (s)", y = "Delta Energy Consumption (Joules)") +
      theme_minimal()

mixed_only_workload_delta <- mixed_workload_delta[ mixed_workload_delta$work == "ola-mixed",]
single_run_time_model <- glm( seconds ~ dimension*population_size*max_gens, data = mixed_only_workload_delta)
mixed_only_workload_delta$residual_seconds <- residuals( single_run_time_model)

mixed_only_workload_delta$dimension <- as.factor(mixed_only_workload_delta$dimension)
mixed_only_workload_delta$population_size <- as.factor(mixed_only_workload_delta$population_size)
single_run_PKG_model <- glm( PKG ~ residual_seconds + I(residual_seconds^2) + dimension*population_size, data=mixed_only_workload_delta)
```

We some more differentiation still, a few runs occupying the "negative" area, but there is still some confusion and little differentiation for different parameters. 

**Takeaway:** It's a bit better, and it leaves room for more variance explained by changes in the algorithm or its parameters. But the main issue is that system state is not *totally* eliminated by subtracting baseline.

# Reducing Variability through Repetition

Can multiple runs across different system states yield an energy profile that allows us to compare between different algorithm configurations and representations?

Let us try and repeat the experiments *five* times; measurements of energy for every parameter configuration are then repeated *30 x 5* times. We will be also using a *sandwich* strategy: every workload experiment will be *sandwiched* between two baseline experiments.

```{r multi-run-results, fig.cap="Energy consumption still depends on time"}
files <- c("data/cec-1.12.4-25.10-cec-mixed-v1-26-Jan-08-30-26.csv",
           "data/cec-1.12.4-25.10-cec-mixed-v2-26-Jan-12-20-59.csv",
           "data/cec-1.12.4-25.10-cec-mixed-v3-26-Jan-17-42-27.csv",
           "data/cec-1.12.4-25.10-cec-mixed-v4-27-Jan-08-09-20.csv",
           "data/cec-1.12.4-25.10-cec-mixed-v5-27-Jan-10-16-05.csv")

all_data <- do.call(rbind, lapply(files, function(f) process_deltas(read.csv(f))))

all_data$dimension <- as.factor( all_data$dimension)
all_data$population_size <- as.factor( all_data$population_size)
all_data$max_gens <- as.factor(all_data$max_gens)

multi_run_time_model <- glm( seconds ~ dimension*population_size*max_gens, data = all_data)
all_data$residual_seconds <- residuals( multi_run_time_model)

multi_run_PKG_model <- glm( PKG ~ residual_seconds + I(residual_seconds^2) + dimension*population_size*max_gens, data=all_data)

library(marginaleffects)

# ---------------------------------------------------------
# GRAPH 1: The Time Effect (Quadratic)
# ---------------------------------------------------------
library(ggplot2)
library(marginaleffects)

pred_tiempo <- predictions(multi_run_PKG_model, 
                           newdata = datagrid(residual_seconds = seq(-1, 4.8, by = 0.2)))

# 2. Dibujar el gráfico superponiendo la predicción y los datos brutos
ggplot(pred_tiempo, aes(x = residual_seconds, y = estimate)) +
  
  # Puntos de datos crudos
  # (Nota: asumo que en all_data la columna se llama residual_seconds. 
  # Si sigue siendo 'seconds', cambia el x= de aquí abajo a x=seconds)
  geom_point(data = all_data, 
             aes(x = residual_seconds, y = PKG, color = interaction(dimension, population_size)), 
             alpha = 0.3) + 
  
  # Banda del intervalo de confianza del modelo
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "blue", alpha = 0.4) +
  
  # Línea predictiva del modelo
  geom_line(color = "darkblue", linewidth = 1.2) + 
  
  labs(title = "Marginal time effect in PKG over raw data",
       x = "Time (residual seconds)",
       y = "Energy consumption (PKG)",
       color = "Dimension x Population size") +
  theme_minimal()

# Check how much we have improved
# 1. Extract 95% Confidence Intervals for both models
ci_single <- confint(single_run_PKG_model)
ci_multi  <- confint(multi_run_PKG_model)

# 2. Pick a parameter that exists in both to compare (e.g., population_size)
# Replace 'population_size' with the exact name of the row from your summary
param <- "dimension5:population_size400"

# 3. Calculate the Width (Upper Bound - Lower Bound)
width_single <- ci_single[param, 2] - ci_single[param, 1]
width_multi  <- ci_multi[param, 2] - ci_multi[param, 1]

#cat("Uncertainty (CI Width) from 1 Block: ", round(width_single, 2), " PKG units\n")
#cat("Uncertainty (CI Width) from 5 Blocks: ", round(width_multi, 2), " PKG units\n")

# 4. Calculate the ROI of your time investment
certainty_gained <- ((width_single - width_multi) / width_single) * 100
# cat("-> By investing 5x the time, we reduced uncertainty regarding the true effect of population size by", round(certainty_gained, 2), "%\n")
```


This is a plot of the dependence of energy consumption on running time. Running time is important, because it is one of the variables where system state shows; as frequency shifts the workload will take more or less time; that might be a consequence of the system moving workload to lower-frequency cores or a dynamic change in voltage that shifts frequency. 

We have used a two-stage model here. Check [this explanation on the two-stage process](cec-two-stage-explainer.html) by which we first model time to separate its dependence on the algorithm variables from the unexplained variance, and then energy.

The created model (you can check it out by clicking on the show code button) points towards that direction, with a positive coefficient on the `seconds` term and a negative one on the squared seconds term; this is clearly shown in this chart. There is still some confusion between problems of different dimensions and population sizes, but this is also because one of the most significant variables, `max_gens` is not represented. 

**Takeaway**: Since running time is the most important factor in energy consumption, even if an important part is governed by operating conditions, there is still a part we can act on to reduce energy consumption: increase performance by making the algorithm run faster.

But we already knew that. The main thing is that by making 5 experiment blocks, we have reduced uncertainty

```{r confidence.gained}


### Comparing the two models in an absolute way
### 

# 1. Extraer los nombres de los coeficientes de ambos modelos
coefs_single <- rownames(summary(single_run_PKG_model)$coefficients)
coefs_multi  <- rownames(summary(multi_run_PKG_model)$coefficients)

# 2. Identificar qué términos existen en AMBOS modelos (para una comparación justa)
terminos_compartidos <- intersect(coefs_single, coefs_multi)

# 3. Extraer los Errores Estándar (Std. Error) solo de esos términos compartidos
se_single_shared <- summary(single_run_PKG_model)$coefficients[terminos_compartidos, "Std. Error"]
se_multi_shared  <- summary(multi_run_PKG_model)$coefficients[terminos_compartidos, "Std. Error"]

# 4. Calcular el Error Estándar Medio (Incertidumbre Global Promedio)
incertidumbre_global_single <- mean(se_single_shared)
incertidumbre_global_multi  <- mean(se_multi_shared)

# 5. Calcular la Varianza Total de los Parámetros (A-Optimalidad)
# Esta es la suma de las varianzas (Std. Error al cuadrado)
a_opt_single <- sum(se_single_shared^2)
a_opt_multi  <- sum(se_multi_shared^2)

# 6. Calcular el Retorno de Inversión (ROI) en Porcentaje
mejora_incertidumbre <- ((incertidumbre_global_single - incertidumbre_global_multi) / incertidumbre_global_single) * 100
mejora_varianza <- ((a_opt_single - a_opt_multi) / a_opt_single) * 100

```

With a single experiment block, the precision we could achieve was `r round(incertidumbre_global_single, 2)` Joules. No change under that could be detected. With 5 blocks, we have boosted it to `r round(incertidumbre_global_multi, 2)` Joules. Even if the difference between the energy spent by two values of the parameter is this small, we'll be able to detect it.

We can look at it from another point of view, looking at the total variance of coefficients. From the 1-block model we used before to this one we have achieved an improvement of `r round(mejora_varianza, 2)` %.

**Takeaway**: Since the operating context is unstable, a single experiment block is not enough to achieve enough precision to detect the impact of different variable values on energy. We need several experiment blocks.


# Analyzing Energy Distribution

How does the distribution of delta energy differ across configurations?

```{r ridgeline-delta, fig.cap="Distribution of delta energy for different configurations."}
library(ggridges)
all_data$config <- paste(all_data$dimension, "D, P=", all_data$population_size)
ggplot(all_data, aes(x = delta_PKG, y = config, fill = config)) +
  geom_density_ridges() +
  labs(x = "Delta Energy (Joules)", y = "Configuration") +
  theme_minimal()
```

The main difference is how different population sizes fall across the two "energy peaks". In the cases where energy distribution fall across two peaks, we will see higher energy consumption. The peaks, however, are slightly displaced with respect to each other to this higher precision achieved gives us a bit of leeway to choose among different configurations for the same problem size, or to distinguish among different algorithm implementations.

**Takeaway:** Since we have a good amount of data, we can distinguish now between relatively small energy deltas, giving us more leverage to configure our algorithms in a *greener* way.


# Practical Implications

This methodology allows researchers to treat energy consumption as a first-class optimization objective. By identifying that smaller populations and shorter stagnation limits (max_gens = 10) consume significantly less energy across all tested dimensions, we can make informed trade-offs between solution quality and carbon footprint.

# Reproducibility Notes
All experiments were conducted on an AMD Ryzen 9 9950X system using the `pinpoint` tool for RAPL access. The analysis code and datasets are available in the project repository [@merelo2022brave].

# References
