---
title: "Unmasking Algorithmic Energy: A Two-Stage GLM Methodology"
author: "Methodology Explainer"
date: today
format: 
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: show
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
---

## 1. The Challenge: Hardware Noise vs. Algorithmic Truth

When measuring the energy consumption of an evolutionary algorithm, the raw measurements (`PKG` or `delta_PKG`) are heavily contaminated by the operating environment. A single execution's energy footprint is determined by two competing forces:

1. **The Algorithmic Intent:** The theoretical cost demanded by the problem parameters (Dimension, Population Size, Max Generations).
2. **The Environmental Friction:** Operating system background tasks, CPU frequency scaling, and cache misses that occur randomly during execution.

If we attempt to model energy directly against time and parameters simultaneously, we run into **multicollinearity**—the algorithm's parameters inherently dictate the execution time, which confuses the statistical model. 

To solve this, we employ a **Two-Stage Generalized Linear Model (GLM)** to surgically separate the environmental noise from the pure algorithmic cost.

### 1.1 Data Loading and Preparation

First, we load our 5 independent experimental blocks. By relying on the `process_deltas()` function from the `energyR` package, we ensure that static baseline power is subtracted, leaving us with purely dynamic execution metrics (the Sandwich Protocol).

```{r}
#| label: load-data

library(ggplot2)
library(dplyr)
library(marginaleffects)

# Load the Sandwich protocol delta processing logic
source("energyR/R/process_deltas.R")

# Load the 5 experimental blocks
files <- c("data/cec-1.12.4-25.10-cec-mixed-v1-26-Jan-08-30-26.csv",
           "data/cec-1.12.4-25.10-cec-mixed-v2-26-Jan-12-20-59.csv",
           "data/cec-1.12.4-25.10-cec-mixed-v3-26-Jan-17-42-27.csv",
           "data/cec-1.12.4-25.10-cec-mixed-v4-27-Jan-08-09-20.csv",
           "data/cec-1.12.4-25.10-cec-mixed-v5-27-Jan-10-16-05.csv")

# Combine and process the deltas
all_data <- do.call(rbind, lapply(files, function(f) process_deltas(read.csv(f))))

# Ensure parameters are treated as discrete categorical factors
all_data$dimension <- as.factor(all_data$dimension)
all_data$population_size <- as.factor(all_data$population_size)
all_data$max_gens <- as.factor(all_data$max_gens)
```

### 1.2 The Collinearity Trap (Why a single model fails)

Before applying the two-stage method, let's visualize why time and parameters cannot share the same predictive space without adjustments.

```{r}
#| label: plot-collinearity
#| fig-cap: "Severe correlation between execution time and problem dimension."

ggplot(all_data, aes(x = seconds, y = PKG, color = dimension)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  labs(
    title = "The Confounding Effect: Time vs. Energy",
    subtitle = "Larger dimensions inherently take more time, tangling the variables.",
    x = "Execution Time (Seconds / Delta Seconds)",
    y = "Energy Consumption (PKG / Delta PKG)",
    color = "Problem\nDimension"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

Because `dimension` fundamentally drives `seconds`, a single model cannot determine if an energy spike is caused by the geometry of the problem or simply because the clock ticked longer.

---

## 2. Stage 1: The Time Model (Extracting the Context)

To break the collinearity, we first model the algorithm's execution time based *exclusively* on its configuration parameters.

```{r}
#| label: stage1-time-model

# Stage 1: How LONG SHOULD the algorithm take based on its parameters?
multi_run_time_model <- glm(seconds ~ dimension * population_size * max_gens, data = all_data)

# Extract the residuals: The difference between expected time and actual time
all_data$residual_seconds <- residuals(multi_run_time_model)
```

### Understanding the Residuals
By extracting the `residual_seconds`, we have quantified the "Environmental Friction":
* **`Residual = 0`**: The operating system executed the algorithm exactly in the expected time.
* **`Residual > 0`**: The server was "noisy" or busy, causing the execution to drag on longer than the pure parameters required.
* **`Residual < 0`**: The server executed the task unusually fast (e.g., optimal cache hits).

Let's visualize the shape of this environmental noise across our 5 experimental blocks.

```{r}
#| label: plot-residuals
#| fig-cap: "Distribution of Operating System noise isolated from algorithmic complexity."

ggplot(all_data, aes(x = residual_seconds)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = "darkblue", linewidth = 1) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 0.5, y = 0.1, label = "Slower than expected\n(OS Friction)", color = "red", hjust = 0) +
  annotate("text", x = -0.5, y = 0.1, label = "Faster than expected", color = "darkgreen", hjust = 1) +
  labs(
    title = "Isolated Environmental Noise (Residual Seconds)",
    subtitle = "This metric is mathematically entirely orthogonal to our parameters.",
    x = "Residual Time (Seconds)",
    y = "Density"
  ) +
  theme_minimal()
```

---

## 3. Stage 2: The Energy Model (The Clean Footprint)

With the OS noise mathematically isolated (`residual_seconds`), we can now build our definitive energy model. Because the residuals are completely uncorrelated with the `dimension` or `population_size` variables, multicollinearity is reduced to zero.

```{r}
#| label: stage2-pkg-model

# Stage 2: Modeling Energy using isolated noise + pure algorithmic parameters
multi_run_PKG_model <- glm(PKG ~ residual_seconds + I(residual_seconds^2) + 
                                 dimension * population_size * max_gens, 
                           data = all_data)
```

### 3.1 Visualizing the Hardware Penalty
First, let's look at what the `residual_seconds` terms are doing. They act as a mathematical sponge, absorbing the non-linear energy penalty of a busy operating system.

```{r}
#| label: plot-penalty
#| fig-cap: "The isolated effect of environmental friction on total energy consumption."

# Generate predictions varying ONLY the noise, keeping base parameters constant
pred_noise <- predictions(
  multi_run_PKG_model, 
  newdata = datagrid(residual_seconds = seq(min(all_data$residual_seconds), 
                                            max(all_data$residual_seconds), 
                                            length.out = 100))
)

ggplot(pred_noise, aes(x = residual_seconds, y = estimate)) +
  geom_point(data = all_data, aes(x = residual_seconds, y = PKG), 
             color = "grey", alpha = 0.2, size = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "firebrick", alpha = 0.3) +
  geom_line(color = "darkred", linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "black", linewidth = 1) +
  labs(
    title = "The Energy Cost of Operating System Friction",
    subtitle = "The quadratic curve absorbs the hardware penalty, cleaning the parameters.",
    x = "Environmental Noise (Residual Seconds)",
    y = "Predicted Energy (PKG)"
  ) +
  theme_minimal()
```

### 3.2 Visualizing the Pure Algorithmic Truth
Finally, because the Stage 2 model has absorbed the hardware noise, the coefficients for our parameters (`dimension`, `population_size`, `max_gens`) now represent the **pure algorithmic truth**. 

We can ask the model: *"What would the energy consumption be if the operating system were perfectly silent (`residual_seconds = 0`)?"*

```{r}
#| label: plot-truth
#| fig-cap: "The pure, unconfounded energy cost of evolutionary parameters."

# Predict energy across different dimensions and populations, FORCING noise to 0
pred_algo <- predictions(
  multi_run_PKG_model,
  newdata = datagrid(
    dimension = unique(all_data$dimension),
    population_size = unique(all_data$population_size),
    max_gens = unique(all_data$max_gens)[1], # Keep max_gens constant for 2D visualization
    residual_seconds = 0                     # The crucial step: silencing the hardware noise
  )
)

ggplot(pred_algo, aes(x = dimension, y = estimate, color = population_size, group = population_size)) +
  geom_point(size = 4) +
  geom_line(linewidth = 1.2) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, linewidth = 0.8) +
  labs(
    title = "Unmasked Algorithmic Energy Footprint",
    subtitle = "Modeled under 'perfect' OS conditions (Residual = 0) revealing the true interaction.",
    x = "Problem Dimension",
    y = "Predicted Pure Energy Cost (PKG)",
    color = "Population\nSize"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## 4. Conclusion

By utilizing this two-stage methodology alongside 5 blocks of experimental replications:
1. We **avoided the multicollinearity trap** that invalidates standard single-stage energy models.
2. We **isolated the macro-noise** of the hardware environment into an orthogonal vector (`residual_seconds`).
3. We secured **narrow confidence intervals** for complex 3-way interactions, allowing us to definitively prove how algorithm geometry and search density combine to consume physical server energy.

---

## References

* **Gelman, A., & Hill, J. (2006).** *Data Analysis Using Regression and Multilevel/Hierarchical Models* (1st ed.). Cambridge University Press. DOI: [10.1017/CBO9780511790942](https://doi.org/10.1017/CBO9780511790942)
* **McCullagh, P., & Nelder, J. A. (1989).** *Generalized Linear Models* (2nd ed.). Chapman and Hall/CRC. DOI: [10.1201/9780203753736](https://doi.org/10.1201/9780203753736)
