---
title: "LION 2026 in 4 Visuals"
subtitle: "Energy and fitness trade-offs in the Brave New Algorithm"
author: "JJ Merelo & Cecilia Merelo-Molina"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
    fig_caption: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5,
  dpi = 150
)
```

```{r libraries}
library(ggplot2)
library(dplyr)

if (!requireNamespace("energyR", quietly = TRUE)) {
  if (file.exists("energyR/DESCRIPTION")) {
    utils::install.packages("energyR", repos = NULL, type = "source")
  } else {
    stop("Package 'energyR' is required. Install it or render from the repository root.")
  }
}

library(energyR)
```

## Why this explainer?

The LION paper evaluates differential evolution (DE), a population-based black-box optimization algorithm, on BBOB numerical optimization tasks while jointly tracking fitness and energy.

In this context, **baseline** refers to the per-configuration reference measurements used to estimate non-workload energy before computing workload deltas. This page condenses the core paper findings into four visual messages:

1. Baseline energy is noisy and must be discounted carefully.
2. Within the same Julia version, `max_gens = 10` usually saves energy.
3. Lower energy and better fitness do not always happen together.
4. Upgrading Julia can change the entire energy profile, even with the same algorithm.

## Data preparation with `energyR`

```{r data-prep}
# LION (Julia 1.11.8)
lion_baseline <- load_bna_csv("data/lion-1.11.8-baseline.csv")
lion_summary <- summarize_baseline(lion_baseline)

lion_workload <- load_bna_csv("data/lion-1.11.8-bna-fix-rand.csv", drop_baseline_cols = FALSE)
lion_workload <- prepare_workload(lion_summary, lion_workload, label_col = "dim_pop")
lion_workload$julia <- "1.11.8"

# EvoApps baseline/workload used in the paper as the previous platform (Julia 1.11.7)
prev_baseline <- load_bna_csv("data/evoapps-1.11.7-baseline-bna-baseline-16-Oct-11-08-20.csv")
prev_summary <- summarize_baseline(prev_baseline)

prev_workload <- load_bna_csv("data/evoapps-1.11.7-fix-rand-bna-fix-rand-25-Oct-11-06-07.csv", drop_baseline_cols = FALSE)
prev_workload <- prepare_workload(prev_summary, prev_workload, label_col = "dim_pop")
prev_workload$julia <- "1.11.7"
```

## 1) Baseline is variable, so subtraction matters

Before looking at algorithmic effects, we need to separate them from runtime and platform overhead. The baseline plot shows that baseline measurements have non-negligible dispersion, and that dispersion depends on dimension and population size.

This is one of the main methodological conclusions of the paper: comparing raw Joules between configurations is misleading. Baseline-corrected deltas are not an optional refinement; they are the measurement unit that makes cross-setting comparisons defensible.

```{r baseline-variability, fig.cap="Baseline PKG distribution by dimension and population size (LION, Julia 1.11.8)."}
plot_data <- lion_baseline %>%
  mutate(
    dimension = as.factor(dimension),
    population_size = as.factor(population_size)
  )

ggplot(plot_data, aes(x = dimension, y = PKG, fill = population_size)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.15, alpha = 0.8, outlier.alpha = 0.2) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Baseline runs have visible spread",
    subtitle = "This is why LION computes delta energy against per-configuration baseline summaries",
    x = "Dimension",
    y = "PKG energy (J)",
    fill = "Population"
  )
```

## 2) Main LION configuration effect: stopping earlier usually saves energy

With baseline correction in place, the effect of `max_gens` becomes clearer: in most parameter combinations, `max_gens = 10` shifts the distribution downward relative to larger values.

The practical conclusion is not "always use fewer generations." The paper shows that this setting is an energy-lean default, but it must be interpreted jointly with solution quality and with problem-specific accuracy requirements.

```{r lion-energy-by-maxgens, fig.cap="Delta energy in LION workload runs, grouped by max_gens and faceted by parameter combination."}
lion_for_plot <- lion_workload %>%
  mutate(
    work = paste0("max_gens=", max_gens),
    dim_pop = factor(dim_pop)
  )

plot_delta_energy(
  lion_for_plot,
  geom = "boxplot",
  facet_col = "dim_pop",
  title = "Delta energy by stopping criterion (LION / Julia 1.11.8)"
)
```

## 3) Energy vs fitness is a trade-off, not a single objective

This plot makes explicit what the paper discusses in detail: there is no universal setting that simultaneously optimises energy and fitness in every case. Points do not collapse into a single dominant frontier.

Operationally, that means configuration should be policy-driven. If your context prioritises lower energy, choose from the lower-energy clusters; if it prioritises better fitness improvements, accept the additional energy budget and select accordingly.

```{r energy-fitness-tradeoff, fig.cap="Energy-versus-fitness view for LION runs. Lower-left is ideal, but not all settings move there together."}
lion_tradeoff <- lion_for_plot %>%
  filter(is.finite(log_diff), diff_fitness > 0) %>%
  mutate(work = paste0("max_gens=", max_gens))

plot_energy_vs_fitness(
  lion_tradeoff,
  x_col = "delta_PKG",
  y_col = "log_diff",
  color_col = "work",
  alpha = 0.55,
  title = "LION: energy vs fitness difference"
)
```

## 4) Platform drift: Julia version alone changes the energy baseline-corrected profile

One of the strongest discussion points in the paper is reproducibility under software platform evolution. The comparison across Julia versions shows that runtime/platform updates can materially shift measured energy, even when the algorithmic procedure is unchanged.

So, any claim of "configuration A is greener than B" should be version-scoped and re-validated after upgrades. In practice, rerunning baseline + workload after runtime changes is part of the protocol, not a nice-to-have.

```{r julia-version-compare, fig.cap="Comparing baseline-corrected workload energy for similar BNA settings across Julia versions used in the paper."}
version_compare <- bind_rows(
  prev_workload %>% mutate(source = "EvoApps setup"),
  lion_workload %>% mutate(source = "LION setup")
) %>%
  mutate(
    population_size = as.factor(population_size),
    dimension = as.factor(dimension)
  )

ggplot(version_compare, aes(x = julia, y = delta_PKG, fill = julia)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  facet_grid(dimension ~ population_size) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Julia version impacts observed energy",
    subtitle = "Consistent with the LION paper: software platform changes can dominate small tuning effects",
    x = "Julia version",
    y = "Delta PKG energy (J)",
    fill = "Julia"
  )
```

## Quick scoreboard for a social-media thread

```{r summary-table}
lion_summary_table <- create_summary(lion_workload) %>%
  arrange(trimmed_mean_delta_PKG) %>%
  mutate(
    setting = paste0("D=", dimension, ", P=", population_size, ", max_gens=", max_gens),
    trimmed_mean_delta_PKG = round(trimmed_mean_delta_PKG, 2),
    trimmed_mean_energy_per_evaluation = signif(trimmed_mean_energy_per_evaluation, 3)
  ) %>%
  select(setting, trimmed_mean_delta_PKG, trimmed_mean_energy_per_evaluation)

knitr::kable(
  lion_summary_table,
  col.names = c("Setting", "Trimmed mean delta PKG (J)", "Trimmed mean J/evaluation"),
  caption = "Lower values are better for energy-only efficiency."
)
```

The table is useful as an at-a-glance ranking, but it should be read as a screening tool. Final decisions should combine this with fitness outcomes and with the platform/version context highlighted above.

## Discussion highlights from the paper

- **Measurement methodology matters as much as algorithm design:** baseline correction and careful sampling choices directly influence the conclusions you can trust.
- **Energy is a first-class optimisation objective:** if it is not explicitly included in analysis, tuning decisions can look good on fitness while being unnecessarily expensive energetically.
- **Results are contextual, not absolute:** hardware state, runtime version, and execution environment can dominate small parameter effects.
- **Reproducibility requires protocol, not just code:** to compare studies over time, rerun baseline and workload suites under each relevant software/hardware change.
- **Actionable takeaway:** report both fitness and energy, and present trade-off-aware recommendations instead of one-number "best setting" claims.

## Suggested 5-post thread structure

1. **Context:** "We profiled BNA energy in LION with baseline-corrected measurements, not raw Joules."
2. **Method:** "Baseline spread is real, so using per-configuration baseline summaries is mandatory."
3. **Result:** "Lower `max_gens` often reduces the amount of energy spent for a certain configuration, but can limit final fitness."
4. **Trade-off:** "The best setting depends on whether your priority is raw energy savings or better fitness."
5. **Platform warning:** "Runtime upgrades can shift energy behaviour by themselves; always re-measure after updates."
