---
title: "LION 2026 in 4 Visuals"
subtitle: "Energy and fitness trade-offs in the Brave New Algorithm"
author: "JJ Merelo & Cecilia Merelo-Molina"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
    fig_caption: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5,
  dpi = 150
)
```

```{r libraries}
library(ggplot2)
library(dplyr)

if (requireNamespace("energyR", quietly = TRUE)) {
  summarize_baseline <- energyR::summarize_baseline
  compute_deltas <- energyR::compute_deltas
  create_summary <- energyR::create_summary
  add_log_diff <- energyR::add_log_diff
  add_pop_dim_label <- energyR::add_pop_dim_label
  plot_delta_energy <- energyR::plot_delta_energy
  plot_energy_vs_fitness <- energyR::plot_energy_vs_fitness
  load_bna_csv <- energyR::load_bna_csv
} else {
  source("energyR/R/summarize_baseline.R")
  source("energyR/R/compute_deltas.R")
  source("energyR/R/create_summary.R")
  source("energyR/R/add_log_diff.R")
  source("energyR/R/add_pop_dim_label.R")
  source("energyR/R/plot_delta_energy.R")
  source("energyR/R/plot_energy_vs_fitness.R")
  source("energyR/R/load_bna_csv.R")
}
```

## Why this explainer?

The LION paper studies one practical question: can we improve algorithmic fitness without paying too much extra energy?

To make the main findings easier to share in a short blog post or thread, this page condenses the full paper into four visual messages:

1. Baseline energy is noisy and must be discounted carefully.
2. Within the same Julia version, `max_gens = 10` usually saves energy.
3. Lower energy and better fitness do not always happen together.
4. Upgrading Julia can change the entire energy profile, even with the same algorithm.

## Data preparation with `energyR`

```{r data-prep}
# LION (Julia 1.11.8)
lion_baseline <- load_bna_csv("data/lion-1.11.8-baseline.csv")
lion_summary <- summarize_baseline(lion_baseline)

lion_workload <- load_bna_csv("data/lion-1.11.8-bna-fix-rand.csv", drop_baseline_cols = FALSE)
lion_workload <- compute_deltas(lion_summary, lion_workload)
lion_workload <- add_log_diff(lion_workload)
lion_workload <- add_pop_dim_label(lion_workload, label_col = "dim_pop")
lion_workload$julia <- "1.11.8"

# EvoApps baseline/workload used in the paper as the previous platform (Julia 1.11.7)
prev_baseline <- load_bna_csv("data/evoapps-1.11.7-baseline-bna-baseline-16-Oct-11-08-20.csv")
prev_summary <- summarize_baseline(prev_baseline)

prev_workload <- load_bna_csv("data/evoapps-1.11.7-fix-rand-bna-fix-rand-25-Oct-11-06-07.csv", drop_baseline_cols = FALSE)
prev_workload <- compute_deltas(prev_summary, prev_workload)
prev_workload <- add_log_diff(prev_workload)
prev_workload <- add_pop_dim_label(prev_workload, label_col = "dim_pop")
prev_workload$julia <- "1.11.7"
```

## 1) Baseline is variable, so subtraction matters

```{r baseline-variability, fig.cap="Baseline PKG distribution by dimension and population size (LION, Julia 1.11.8)."}
plot_data <- lion_baseline %>%
  mutate(
    dimension = as.factor(dimension),
    population_size = as.factor(population_size)
  )

ggplot(plot_data, aes(x = dimension, y = PKG, fill = population_size)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.15, alpha = 0.8, outlier.alpha = 0.2) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Baseline runs have visible spread",
    subtitle = "This is why LION computes delta energy against per-configuration baseline summaries",
    x = "Dimension",
    y = "PKG energy (J)",
    fill = "Population"
  )
```

## 2) Main LION configuration effect: stopping earlier usually saves energy

```{r lion-energy-by-maxgens, fig.cap="Delta energy in LION workload runs, grouped by max_gens and faceted by parameter combination."}
lion_for_plot <- lion_workload %>%
  mutate(
    work = paste0("max_gens=", max_gens),
    dim_pop = factor(dim_pop)
  )

plot_delta_energy(
  lion_for_plot,
  geom = "boxplot",
  facet_col = "dim_pop",
  title = "Delta energy by stopping criterion (LION / Julia 1.11.8)"
)
```

## 3) Energy vs fitness is a trade-off, not a single objective

```{r energy-fitness-tradeoff, fig.cap="Energy-versus-fitness view for LION runs. Lower-left is ideal, but not all settings move there together."}
lion_tradeoff <- lion_for_plot %>%
  filter(is.finite(log_diff), diff_fitness > 0) %>%
  mutate(work = paste0("max_gens=", max_gens))

plot_energy_vs_fitness(
  lion_tradeoff,
  x_col = "delta_PKG",
  y_col = "log_diff",
  color_col = "work",
  alpha = 0.55,
  title = "LION: energy vs fitness difference"
)
```

## 4) Platform drift: Julia version alone changes the energy baseline-corrected profile

```{r julia-version-compare, fig.cap="Comparing baseline-corrected workload energy for similar BNA settings across Julia versions used in the paper."}
version_compare <- bind_rows(
  prev_workload %>% mutate(source = "EvoApps setup"),
  lion_workload %>% mutate(source = "LION setup")
) %>%
  mutate(
    population_size = as.factor(population_size),
    dimension = as.factor(dimension)
  )

ggplot(version_compare, aes(x = julia, y = delta_PKG, fill = julia)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  facet_grid(dimension ~ population_size) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Julia version impacts observed energy",
    subtitle = "Consistent with the LION paper: software platform changes can dominate small tuning effects",
    x = "Julia version",
    y = "Delta PKG energy (J)",
    fill = "Julia"
  )
```

## Quick scoreboard for a social-media thread

```{r summary-table}
lion_summary_table <- create_summary(lion_workload) %>%
  arrange(trimmed_mean_delta_PKG) %>%
  mutate(
    setting = paste0("D=", dimension, ", P=", population_size, ", max_gens=", max_gens),
    trimmed_mean_delta_PKG = round(trimmed_mean_delta_PKG, 2),
    trimmed_mean_energy_per_evaluation = signif(trimmed_mean_energy_per_evaluation, 3)
  ) %>%
  select(setting, trimmed_mean_delta_PKG, trimmed_mean_energy_per_evaluation)

knitr::kable(
  lion_summary_table,
  col.names = c("Setting", "Trimmed mean delta PKG (J)", "Trimmed mean J/evaluation"),
  caption = "Lower values are better for energy-only efficiency."
)
```

## Suggested 5-post thread structure

1. **Context:** "We profiled BNA energy in LION with baseline-corrected measurements, not raw Joules."
2. **Method:** "Baseline spread is real, so using per-configuration baseline summaries is mandatory."
3. **Result:** "Lower `max_gens` often reduces energy, but can limit final fitness."
4. **Trade-off:** "The best setting depends on whether your priority is raw energy savings or better fitness."
5. **Platform warning:** "Runtime upgrades can shift energy behaviour by themselves; always re-measure after updates."
