---
title: "LION 2026 in 4 Visuals"
subtitle: "Energy and fitness trade-offs in the Brave New Algorithm"
author: "JJ Merelo & Cecilia Merelo-Molina & Pablo García Sánchez & Mario García Valdez & Juan Luis Jiménez-Laredo"
date: "`r format(Sys.Date(), '%B %Y')`"
bibliography:
  - references.bib
  - ours.bib
  - GAs.bib
  - ga-energy.bib
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
    fig_caption: true
    code_folding: hide
---

```{r setup_and_libraries, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5,
  dpi = 150
)

library(ggplot2)
library(dplyr)

if (!requireNamespace("energyR", quietly = TRUE)) {
  if (file.exists("energyR/DESCRIPTION")) {
    pkg_path <- normalizePath("energyR", mustWork = TRUE)
    utils::install.packages(pkg_path, repos = NULL, type = "source")
  } else {
    stop("Package 'energyR' is required. Install it or render from the repository root.")
  }
}

library(energyR)
```

## Introduction

The LION paper analyzes the Brave New Algorithm (BNA), a caste-stratified population-based stochastic optimization algorithm, measuring fitness and energy jointly [@lion26; @merelo2022brave].
The optimization task in this study is the BBOB Sphere function [@hansen2010comparing].

In this context, **baseline** refers to runs that stop after generating the initial population (same code path, no evolutionary loop), computed per `(dimension, population)` configuration and subtracted from workload runs [@lion26]. This page condenses the core paper findings into four visual messages:

1. Baseline energy is noisy and must be discounted carefully.
2. Within the same Julia version, `max_gens = 10` usually saves energy.
3. Lower energy and better fitness do not always happen together.
4. Upgrading Julia can change the entire energy profile, even with the same algorithm.

## Data preparation with `energyR`

```{r data-prep}
# LION (Julia 1.11.8)
lion_baseline <- load_bna_csv("data/lion-1.11.8-baseline.csv")
lion_summary <- summarize_baseline(lion_baseline)

lion_workload <- load_bna_csv("data/lion-1.11.8-bna-fix-rand.csv", drop_baseline_cols = FALSE)
lion_workload <- prepare_workload(lion_summary, lion_workload, label_col = "dim_pop")
lion_workload$julia <- "1.11.8"

# EvoApps baseline/workload used in the paper as the previous platform (Julia 1.11.7)
prev_baseline <- load_bna_csv("data/evoapps-1.11.7-baseline-bna-baseline-16-Oct-11-08-20.csv")
prev_summary <- summarize_baseline(prev_baseline)

prev_workload <- load_bna_csv("data/evoapps-1.11.7-fix-rand-bna-fix-rand-25-Oct-11-06-07.csv", drop_baseline_cols = FALSE)
prev_workload <- prepare_workload(prev_summary, prev_workload, label_col = "dim_pop")
prev_workload$julia <- "1.11.7"
```

## 1) Baseline is variable, so subtraction matters

Before looking at algorithmic effects, we need to separate them from runtime and platform overhead. The baseline plot shows non-negligible dispersion, and that dispersion depends on dimension and population size.

This is one of the main methodological conclusions of the paper: comparing raw Joules between configurations is misleading, so baseline-corrected deltas are the defensible unit for comparisons [@lion26; @cotta24].

```{r baseline-variability, fig.cap="Baseline PKG distribution by dimension and population size (LION, Julia 1.11.8)."}
plot_data <- lion_baseline %>%
  mutate(
    dimension = as.factor(dimension),
    population_size = as.factor(population_size)
  )

ggplot(plot_data, aes(x = dimension, y = PKG, fill = population_size)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Baseline runs have visible spread",
    subtitle = "This is why LION computes delta energy against per-configuration baseline summaries",
    x = "Dimension",
    y = "PKG energy (J)",
    fill = "Population"
  )
```

## 2) Main LION configuration effect: stopping earlier usually saves energy

With baseline correction in place, the effect of `max_gens` becomes clearer: in most parameter combinations, `max_gens = 10` shifts the distribution downward relative to larger values.

The practical conclusion is not "always use fewer generations." The paper shows that this setting is an energy-lean default, but it must be interpreted jointly with solution quality and with problem-specific accuracy requirements.

```{r lion-energy-by-maxgens, fig.cap="Delta energy in LION workload runs, grouped by max_gens and faceted by parameter combination."}
lion_for_plot <- lion_workload %>%
  mutate(
    work = paste0("max_gens=", max_gens),
    dim_pop = factor(dim_pop)
  )

plot_delta_energy(
  lion_for_plot,
  geom = "boxplot",
  facet_col = "dim_pop",
  title = "Delta energy by stopping criterion (LION / Julia 1.11.8)"
)
```

## 3) Energy vs fitness is a trade-off, not a single objective

This plot makes explicit what the paper discusses in detail: there is no universal setting that simultaneously optimizes energy and fitness in every case. Points do not collapse into a single dominant frontier.

Operationally, that means configuration should be policy-driven. If your context prioritizes lower energy, choose from the lower-energy clusters; if it prioritizes better fitness improvements, accept the additional energy budget and select accordingly [@lion26].

```{r energy-fitness-tradeoff, fig.cap="Energy-versus-fitness view for LION runs, faceted by problem dimension and grouped by max_gens/population."}
lion_tradeoff <- lion_for_plot %>%
  filter(is.finite(log_diff), diff_fitness > 0) %>%
  mutate(
    work = factor(paste0("max_gens=", max_gens)),
    dimension = factor(dimension),
    population_size = factor(population_size)
  )

ggplot(lion_tradeoff, aes(x = delta_PKG, y = log_diff, color = work, shape = population_size)) +
  geom_point(alpha = 0.55, size = 2) +
  facet_wrap(~dimension, nrow = 1, scales = "free_x") +
  theme_minimal(base_size = 13) +
  labs(
    title = "LION: energy versus fitness by dimension",
    subtitle = "Faceting by dimension clarifies the per-configuration max_gens effect",
    x = "Delta PKG energy (J)",
    y = "log10 fitness improvement",
    color = "max_gens",
    shape = "Population"
  )
```

## 4) Platform drift: Julia version alone changes the energy baseline-corrected profile

One of the strongest discussion points in the paper is reproducibility under software platform evolution. The comparison across Julia versions shows that runtime/platform updates can materially shift measured energy, even when the algorithmic procedure is unchanged [@lion26].

So, any claim of "configuration A is greener than B" should be version-scoped and re-validated after upgrades. In practice, rerunning baseline + workload after runtime changes is part of the protocol, not a nice-to-have [@cotta24; @lion26].

```{r julia-version-compare, fig.cap="Comparing baseline-corrected workload energy for similar BNA settings across Julia versions used in the paper."}
version_compare <- bind_rows(
  prev_workload %>% mutate(source = "EvoApps setup"),
  lion_workload %>% mutate(source = "LION setup")
) %>%
  mutate(
    population_size = as.factor(population_size),
    dimension = as.factor(dimension)
  )

ggplot(version_compare, aes(x = julia, y = delta_PKG, fill = julia)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  facet_grid(dimension ~ population_size) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Julia version impacts observed energy",
    subtitle = "Consistent with the LION paper: software platform changes can dominate small tuning effects",
    x = "Julia version",
    y = "Delta PKG energy (J)",
    fill = "Julia"
  )
```

## Compact configuration summary

```{r summary-table}
# 1. Create a table of pairwise Wilcoxon tests
pairwise_comparisons <- lion_workload %>%
  filter(is.finite(delta_PKG)) %>%
  group_by(dimension, population_size) %>%
  group_modify(~{
    gens <- sort(unique(.x$max_gens))
    
    # If there's nothing to compare, return an empty template
    if (length(gens) < 2) {
      return(data.frame(
        max_gens_A = numeric(0), 
        max_gens_B = numeric(0), 
        p_value = numeric(0), 
        sig_mark = character(0)
      ))
    }
    
    # Generate all unique pairs of max_gens
    pair_tests <- utils::combn(gens, 2, simplify = FALSE)
    
    # Calculate the test for each pair and keep them in separate columns
    res <- lapply(pair_tests, function(pair) {
      x <- .x$delta_PKG[.x$max_gens == pair[1]]
      y <- .x$delta_PKG[.x$max_gens == pair[2]]
      p_value <- stats::wilcox.test(x, y, exact = FALSE)$p.value
      
      data.frame(
        max_gens_A = pair[1],
        max_gens_B = pair[2],
        p_value = p_value,
        sig_mark = if (is.finite(p_value) && p_value < 0.05) "*" else ""
      )
    })
    
    dplyr::bind_rows(res)
  }) %>%
  ungroup()

# 2. Compute the base summary metrics (assuming create_summary returns one row per config)
base_summary <- create_summary(lion_workload)

# 3. Join the summary metrics for both A and B to the pairwise table
lion_summary_table <- pairwise_comparisons %>%
  # Join metrics for max_gens_A
  left_join(
    base_summary %>% select(dimension, population_size, max_gens, trimmed_mean_delta_PKG, trimmed_mean_energy_per_evaluation),
    by = c("dimension", "population_size", "max_gens_A" = "max_gens")
  ) %>%
  rename(
    delta_PKG_A = trimmed_mean_delta_PKG,
    energy_eval_A = trimmed_mean_energy_per_evaluation
  ) %>%
  # Join metrics for max_gens_B
  left_join(
    base_summary %>% select(dimension, population_size, max_gens, trimmed_mean_delta_PKG, trimmed_mean_energy_per_evaluation),
    by = c("dimension", "population_size", "max_gens_B" = "max_gens")
  ) %>%
  rename(
    delta_PKG_B = trimmed_mean_delta_PKG,
    energy_eval_B = trimmed_mean_energy_per_evaluation
  ) %>%
  # Clean up and arrange
  mutate(
    delta_PKG_A = round(delta_PKG_A, 2),
    delta_PKG_B = round(delta_PKG_B, 2),
    energy_eval_A = signif(energy_eval_A, 3),
    energy_eval_B = signif(energy_eval_B, 3),
    p_value = signif(p_value, 3)
  ) %>%
  # Select final columns in a logical reading order
  select(
    dimension, population_size, 
    max_gens_A, delta_PKG_A, energy_eval_A,
    max_gens_B, delta_PKG_B, energy_eval_B, 
    p_value, sig_mark
  ) %>%
  arrange(dimension, population_size, p_value)

# 4. Render the table
knitr::kable(
  lion_summary_table,
  col.names = c(
    "Dimension", "Pop. Size", 
    "Max Gens A", "Delta PKG A (J)", "J/eval A",
    "Max Gens B", "Delta PKG B (J)", "J/eval B", 
    "p-value", "Sig."
  ),
  caption = "Pairwise comparisons of energy-only efficiency between max_gens settings. Lower values are better. * marks a statistically significant difference (p < 0.05) in delta-PKG between the specific pair in the row."
)

```

The table is useful as an at-a-glance ranking, but it should be read as a screening tool. Final decisions should combine this with fitness outcomes and with the platform/version context highlighted above. We can see, however, that the only significant differences in energy expenses for different values of `max_gens` are in the top rows, with dimension 5 and population 400.

## Expanded analysis: multi-stage model on the final LION dataset

This section applies the later-paper multi-stage modeling workflow directly to the final LION workload file `data/lion-1.11.8-bna-fix-rand.csv` [@lion26; @cotta24]:

1. model runtime (`seconds`) from algorithm covariates (`dimension`, `population_size`, `generations`, `evaluations`);
2. residualize runtime;
3. model energy (`PKG`) on the same covariates plus residualized runtime.

```{r lion-expanded-model-data}
lion_model_data <- read.csv("data/lion-1.11.8-bna-fix-rand.csv") %>%
  mutate(
    max_gens = factor(max_gens),
    alpha = factor(alpha),
    population_size = factor(population_size),
    dimension = factor(dimension)
  )

time_model <- glm(
  seconds ~ dimension*population_size*generations*evaluations,
  data = lion_model_data
)

lion_model_data$residual_seconds <- residuals(time_model)

```
These `residual_seconds` isolate operating-context and implementation influences, not affected by the algorithm parameters.
```{r lion-expanded-energy-model}
energy_model <- lm(
  PKG ~ dimension + population_size + generations + evaluations + residual_seconds,
  data = lion_model_data
)

energy_coef <- summary(energy_model)$coefficients
energy_coef_table <- data.frame(
  term = rownames(energy_coef),
  estimate = energy_coef[, "Estimate"],
  std_error = energy_coef[, "Std. Error"],
  statistic = energy_coef[, "t value"],
  p_value = energy_coef[, "Pr(>|t|)"],
  row.names = NULL
) %>%
  filter(term != "(Intercept)") %>%
  arrange(p_value)

energy_sig_table <- energy_coef_table %>%
  filter(p_value < 0.05) %>%
  mutate(
    estimate = signif(estimate, 4),
    std_error = signif(std_error, 3),
    statistic = signif(statistic, 3),
    p_value = signif(p_value, 3)
  )

if (nrow(energy_sig_table) == 0) {
  energy_sig_table <- data.frame(
    term = "No coefficient with p < 0.05",
    estimate = NA, std_error = NA, statistic = NA, p_value = NA
  )
}

knitr::kable(
  energy_sig_table,
  col.names = c("Term", "Estimate", "Std. error", "t", "p-value"),
  caption = "Stage 2 (energy model): statistically significant coefficients influencing PKG."
)
```

```{r lion-expanded-variance, fig.cap="Variance decomposition for the multi-stage energy model (percent of model sum of squares by term)."}
energy_anova <- anova(energy_model)
energy_variance <- data.frame(
  term = rownames(energy_anova),
  sum_sq = energy_anova$"Sum Sq",
  p_value = energy_anova$"Pr(>F)",
  row.names = NULL
) %>%
  filter(!is.na(sum_sq), term != "Residuals") %>%
  mutate(
    variance_pct = 100 * sum_sq / sum(sum_sq),
    term = factor(term, levels = term[order(variance_pct)])
  )

ggplot(energy_variance, aes(x = term, y = variance_pct, fill = term)) +
  geom_col(width = 0.72, alpha = 0.9) +
  geom_text(
    aes(label = paste0(round(variance_pct, 1), "%")),
    hjust = -0.1,
    size = 3.8
  ) +
  coord_flip() +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  labs(
    title = "Energy variance explained by each modeled term",
    subtitle = "Computed from ANOVA sum-of-squares in the Stage 2 model",
    x = "Model term",
    y = "Share of explained variance (%)"
  )

energy_variance_table <- energy_variance %>%
  mutate(
    variance_pct = round(variance_pct, 2),
    p_value = signif(p_value, 3)
  ) %>%
  arrange(desc(variance_pct)) %>%
  select(term, variance_pct, p_value)

knitr::kable(
  energy_variance_table,
  col.names = c("Term", "Explained variance (%)", "p-value"),
  caption = "ANOVA-based variance share by term for the multi-stage energy model."
)
```

## Discussion highlights from the paper

- **Measurement methodology matters as much as algorithm design:** baseline correction and careful sampling choices directly influence the conclusions you can trust [@lion26; @cotta24].
- **Energy is a first-class optimization objective:** if it is not explicitly included in analysis, tuning decisions can look good on fitness while being unnecessarily expensive energetically [@lion26].
- **Results are contextual, not absolute:** hardware state, runtime version, and execution environment can dominate small parameter effects [@lion26; @cotta24].
- **Reproducibility requires protocol, not just code:** to compare studies over time, rerun baseline and workload suites under each relevant software/hardware change [@lion26].
- **Actionable takeaway:** report both fitness and energy, and present trade-off-aware recommendations instead of one-number "best setting" claims [@lion26; @merelo2022brave].

## References
