---
title: "Workload Energy Measurements vs. Baseline Averages"
subtitle: "Per-parameter-combination breakdown for the OLA-26 experiments"
author: "JJ Merelo, Cecilia Merelo-Molina"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(ggplot2)
library(dplyr)
library(kableExtra)
# ggridges provides ridgeline density plots; install with:
#   install.packages("ggridges")
library(ggridges)

# Colour palette: one colour per (dimension × population-size) combination
combo_palette <- c(
  "Dim 3 | Pop 200" = "#0072B2",
  "Dim 3 | Pop 400" = "#56B4E9",
  "Dim 5 | Pop 200" = "#D55E00",
  "Dim 5 | Pop 400" = "#E69F00"
)
```

# Context

The OLA-26 paper computes *workload energy* (Δ PKG) by subtracting a
**trimmed-mean baseline** from each raw workload measurement.  A table of
baseline statistics is shown but the subtraction step itself is never
visualised: it is impossible to see from the paper how far individual workload
readings sit from their respective baseline average, or whether the spread of
those readings is similar across parameter combinations.

This document fills that gap with three complementary views, all derived from
the 100 ms-sampling baseline (`ola-base-ola-baseline-14-Dec-12-06-42.csv`) and
the matching workload run (`ola-1.11.7-ola-14-Dec-13-02-30.csv`):

1. **Raw workload PKG with per-combination baseline averages overlaid** –
   makes the subtraction step explicit.
2. **Δ PKG distributions by parameter combination** – violin + box-plot.
3. **Ridgeline density of Δ PKG** – alternative density view.

---

# Data preparation

```{r load-data, echo=TRUE}
null_cols <- c("alpha", "max_gens", "different_seeds",
               "diff_fitness", "generations", "evaluations")

# ── Baseline (100 ms sampling) ───────────────────────────────────────────────
baseline <- read.csv("data/ola-base-ola-baseline-14-Dec-12-06-42.csv")
baseline[null_cols] <- NULL

# Per-combination trimmed mean (20 % trimming) used as the subtracted average
baseline_summary <- baseline %>%
  group_by(dimension, population_size) %>%
  summarise(
    n_baseline        = n(),
    trimmed_mean_PKG  = mean(PKG,     trim = 0.2),
    trimmed_mean_secs = mean(seconds, trim = 0.2),
    median_PKG        = median(PKG),
    sd_PKG            = sd(PKG),
    .groups = "drop"
  ) %>%
  mutate(param_combo = paste0("Dim ", dimension, " | Pop ", population_size))

# ── Workload ─────────────────────────────────────────────────────────────────
workload <- read.csv("data/ola-1.11.7-ola-14-Dec-13-02-30.csv")
workload$param_combo <- paste0("Dim ", workload$dimension,
                                " | Pop ", workload$population_size)

# Attach baseline trimmed mean to every workload row (for the overlay line)
workload <- workload %>%
  left_join(
    baseline_summary %>% select(dimension, population_size, trimmed_mean_PKG),
    by = c("dimension", "population_size")
  )

# Compute Δ PKG = workload PKG − trimmed-mean baseline for that combination
workload$delta_PKG <- workload$PKG - workload$trimmed_mean_PKG

# Run index within each parameter combination (x-axis for the first plot)
workload <- workload %>%
  group_by(param_combo) %>%
  mutate(run_index = row_number()) %>%
  ungroup()
```

## Baseline summary table

```{r baseline-table}
kable(
  baseline_summary %>%
    select(param_combo, n_baseline,
           trimmed_mean_PKG, median_PKG, sd_PKG) %>%
    mutate(across(where(is.numeric), ~ round(.x, 2))),
  col.names = c("Parameter combination", "n",
                "Trimmed mean PKG (J)", "Median PKG (J)", "SD (J)"),
  caption = paste("Baseline summary (100 ms sampling).",
                  "The trimmed mean is what is subtracted from each workload run.")
) %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(3, background = "#fff3cd")   # highlight the subtracted value
```

---

# Workload PKG with baseline averages

Each point is **one workload run**.  The dashed line is the **trimmed-mean
baseline for that combination** — the value that will be subtracted.
Vertical segments connect each run to its baseline line:
<span style="color:#009E73">**green**</span> = above baseline (positive Δ PKG),
<span style="color:#CC79A7">**pink**</span> = below baseline (negative Δ PKG).

```{r workload-vs-baseline, fig.cap="Workload PKG energy per run (points) with per-combination trimmed-mean baseline overlaid (dashed line). Vertical segments show the signed Δ PKG.", fig.height=7}
ggplot(workload,
       aes(x = run_index, y = PKG)) +
  # vertical segments: sign-coded
  geom_segment(
    aes(xend = run_index,
        yend = trimmed_mean_PKG,
        colour = delta_PKG >= 0),
    alpha = 0.35, linewidth = 0.5
  ) +
  # workload points
  geom_point(colour = "grey25", size = 2, alpha = 0.75) +
  # baseline average line
  geom_hline(
    data = baseline_summary,
    aes(yintercept = trimmed_mean_PKG),
    linetype = "dashed", colour = "#D55E00", linewidth = 0.9
  ) +
  # baseline ± 1 SD band
  geom_rect(
    data = baseline_summary,
    aes(xmin = -Inf, xmax = Inf,
        ymin = trimmed_mean_PKG - sd_PKG,
        ymax = trimmed_mean_PKG + sd_PKG),
    fill = "#D55E00", alpha = 0.08,
    inherit.aes = FALSE
  ) +
  scale_colour_manual(
    values = c("TRUE" = "#009E73", "FALSE" = "#CC79A7"),
    labels = c("TRUE" = "Above baseline (Δ PKG > 0)",
               "FALSE" = "Below baseline (Δ PKG < 0)"),
    name = NULL
  ) +
  facet_wrap(~ param_combo, ncol = 2, scales = "free_y") +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x     = "Run index within parameter combination",
    y     = "PKG energy (Joules)",
    title = "Workload PKG energy with per-combination baseline average",
    subtitle = paste(
      "Dashed line = trimmed mean of baseline (the subtracted value);",
      "shaded band = ±1 SD of baseline"
    )
  )
```

> **Key observation.** The baseline average (dashed line) is *not* a horizontal
> line that every workload run clusters tightly around: individual workload runs
> can land 100–200 J above or below it, and this spread is fairly consistent
> across all four combinations.  Runs that fall below their own baseline yield
> *negative* Δ PKG — visible here as pink segments — and are not a measurement
> error but an artefact of the separation between baseline and workload
> campaigns.

---

# Δ PKG distributions by parameter combination

## Violin + box-plot

```{r delta-violin, fig.cap="Distributions of Δ PKG (workload PKG − per-combination baseline trimmed mean)", fig.height=5}
ggplot(workload,
       aes(x = param_combo, y = delta_PKG,
           fill = param_combo, colour = param_combo)) +
  geom_violin(alpha = 0.35, linewidth = 0.4, width = 0.85) +
  geom_boxplot(
    width = 0.18, notch = TRUE,
    outlier.size = 1.8, outlier.alpha = 0.5,
    fill = "white", colour = "grey30", alpha = 0.9
  ) +
  geom_hline(yintercept = 0, linetype = "dotted",
             colour = "red", linewidth = 0.8) +
  scale_fill_manual(values = combo_palette)   +
  scale_colour_manual(values = combo_palette) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "none",
    axis.text.x      = element_text(angle = 15, hjust = 1),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x     = "Parameter combination",
    y     = "Δ PKG energy (Joules)",
    title = "Δ PKG = workload PKG − per-combination baseline average",
    subtitle = "Red dotted line at zero · Notch = 95 % CI for the median"
  )
```

## Ridgeline density view

```{r delta-ridge, fig.cap="Ridgeline density of Δ PKG per parameter combination (quantile lines at Q1, median, Q3)", fig.height=4.5}
ggplot(workload,
       aes(x = delta_PKG, y = param_combo,
           fill = param_combo)) +
  geom_density_ridges(
    alpha = 0.55, scale = 0.85,
    quantile_lines = TRUE, quantiles = c(0.25, 0.5, 0.75),
    linewidth = 0.4
  ) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "red", linewidth = 0.8) +
  scale_fill_manual(values = combo_palette) +
  theme_ridges(font_size = 13) +
  theme(legend.position = "none") +
  labs(
    x        = "Δ PKG energy (Joules)",
    y        = NULL,
    title    = "Ridgeline density of Δ PKG per parameter combination",
    subtitle = "Quantile lines at Q1, median, Q3 · Red dashed line at zero"
  )
```

## Summary statistics

```{r delta-summary-table}
delta_summary <- workload %>%
  group_by(param_combo) %>%
  summarise(
    n             = n(),
    median_delta  = round(median(delta_PKG),        2),
    trimmed_mean  = round(mean(delta_PKG, trim = 0.2), 2),
    sd_delta      = round(sd(delta_PKG),            2),
    IQR_delta     = round(IQR(delta_PKG),           2),
    pct_negative  = round(100 * mean(delta_PKG < 0), 1),
    .groups = "drop"
  )

kable(
  delta_summary,
  col.names = c("Parameter combination", "n",
                "Median Δ PKG (J)", "Trimmed mean Δ PKG (J)",
                "SD (J)", "IQR (J)", "% runs below baseline"),
  caption = "Δ PKG summary per parameter combination"
) %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(7,
    background = ifelse(delta_summary$pct_negative > 15,
                        "#f8d7da", "#d4edda"))
```

> **Key observation.** The IQR of Δ PKG is almost as large as the IQR of the
> raw workload PKG (subtracting a *constant* per combination does not reduce
> spread), and the fraction of runs landing below their baseline average ranges
> from `r min(delta_summary$pct_negative)` % to
> `r max(delta_summary$pct_negative)` % depending on the combination.  This
> motivates the sequential *mixed* methodology discussed in the paper, where
> each workload run is paired with the baseline run immediately preceding it,
> so that both capture the system in the same state.

---

# Session information

```{r session-info}
sessionInfo()
```
