---
title: "Getting Started with energyR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with energyR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

## Introduction

The **energyR** package provides reusable utilities for loading, processing,
and visualising energy-consumption data produced by population-based
metaheuristics measured via Intel/AMD RAPL counters.  All helper functions were
distilled from the R code embedded in the papers of the
[Brave New Green Algorithm](https://github.com/JJ/brave-new-green-algorithm)
project; they are general enough to be used in any experiment that follows
the same interleaved-baseline measurement protocol.

This vignette demonstrates the end-to-end workflow using publicly available
data hosted in the project repository.

```{r setup}
library(energyR)
library(ggplot2)
library(dplyr)
```

## 1. Downloading data from the GitHub repository

The package does **not** bundle any data, because the experimental CSV and RDS
files are large.  Instead, we download them on the fly from the GitHub
repository.

```{r download-data}
# Helper: download a file from the project repo if not already cached locally
repo_raw <- "https://raw.githubusercontent.com/JJ/brave-new-green-algorithm/main/data/"

download_bna <- function(filename, destdir = tempdir()) {
  dest <- file.path(destdir, filename)
  if (!file.exists(dest)) {
    url <- paste0(repo_raw, filename)
    tryCatch(
      utils::download.file(url, destfile = dest, quiet = TRUE),
      error = function(e) {
        warning("Could not download ", filename, ": ", conditionMessage(e))
        return(NULL)
      }
    )
  }
  return(dest)
}
```

## 2. Loading baseline data

We use one of the OLA-2026 baseline CSV files as an example.

```{r load-baseline, eval = FALSE}
baseline_file <- download_bna("ola-base-ola-baseline-14-Dec-12-06-42.csv")
baseline_data <- load_bna_csv(baseline_file)   # drops algorithm-parameter cols
head(baseline_data)
```

For the purpose of this vignette we create a small synthetic dataset that
mirrors the structure of the real files so the code runs without network access.

```{r synthetic-baseline}
set.seed(2025)
n_configs <- 4   # 2 dimensions × 2 population sizes
n_reps    <- 30

synthetic_baseline <- do.call(rbind, lapply(seq_len(n_configs), function(i) {
  dim  <- c(3, 3, 5, 5)[i]
  pop  <- c(200, 400, 200, 400)[i]
  data.frame(
    PKG             = rnorm(n_reps, mean = 100 + dim * 2 + pop * 0.05, sd = 5),
    seconds         = rnorm(n_reps, mean = 1.0 + dim * 0.01, sd = 0.05),
    dimension       = dim,
    population_size = pop,
    work            = "BNA-baseline"
  )
}))

head(synthetic_baseline)
```

## 3. Summarising the baseline

```{r summarize-baseline}
baseline_summary <- summarize_baseline(synthetic_baseline)
knitr::kable(baseline_summary, digits = 2,
             caption = "Baseline summary statistics (synthetic data)")
```

## 4. Processing interleaved workload data

When baseline and workload runs are *interleaved* in the same CSV file (one
baseline row, one workload row, one baseline row, ..., grouped in blocks of
61 rows), use `process_deltas()`.

```{r build-interleaved}
# Build a synthetic interleaved data frame (1 block of 61 rows)
block <- data.frame(
  PKG     = rep(NA_real_, 61),
  seconds = rep(NA_real_, 61)
)
# Row 1 = leading baseline
block$PKG[1]     <- rnorm(1, 100, 5)
block$seconds[1] <- rnorm(1, 1.0, 0.05)

for (i in seq(2, 60, by = 2)) {
  # even-index = workload
  block$PKG[i]     <- rnorm(1, 115, 5)
  block$seconds[i] <- rnorm(1, 1.1, 0.05)
  # odd-index = baseline
  block$PKG[i + 1]     <- rnorm(1, 100, 5)
  block$seconds[i + 1] <- rnorm(1, 1.0, 0.05)
}

# Add required grouping columns
block$dimension       <- 3
block$population_size <- 200
block$work            <- ifelse(seq_len(61) %% 2 == 0, "workload", "baseline")

interleaved <- process_deltas(block)
head(interleaved[, c("PKG", "seconds", "delta_PKG", "delta_seconds")])
```

## 5. Computing deltas for a separated-run design

When baseline and workload runs are collected separately, use
`compute_deltas()` together with `summarize_baseline()`.

```{r compute-deltas}
synthetic_workload <- do.call(rbind, lapply(seq_len(n_configs), function(i) {
  dim  <- c(3, 3, 5, 5)[i]
  pop  <- c(200, 400, 200, 400)[i]
  data.frame(
    PKG             = rnorm(n_reps, mean = 115 + dim * 2 + pop * 0.05, sd = 6),
    seconds         = rnorm(n_reps, mean = 1.15 + dim * 0.01, sd = 0.06),
    dimension       = dim,
    population_size = pop,
    max_gens        = 25,
    evaluations     = sample(500:3000, n_reps),
    work            = "BNA-workload"
  )
}))

workload_with_deltas <- compute_deltas(baseline_summary, synthetic_workload)
head(workload_with_deltas[, c("dimension", "population_size",
                               "PKG", "delta_PKG",
                               "seconds", "delta_seconds")])
```

## 6. Summarising workload results

```{r workload-summary}
wl_summary <- create_summary(workload_with_deltas)
knitr::kable(wl_summary, digits = 2,
             caption = "Workload delta-energy summary (synthetic data)")
```

## 7. Statistical tests

`wilcoxon_tests()` compares two work groups for each (dimension,
population_size) combination.

```{r wilcoxon}
combined <- rbind(
  synthetic_baseline[, c("PKG", "seconds", "dimension", "population_size", "work")],
  synthetic_workload[, c("PKG", "seconds", "dimension", "population_size", "work")]
)

wt <- wilcoxon_tests(combined, column_name = "PKG")
knitr::kable(wt, digits = 4,
             caption = "Wilcoxon rank-sum test results (PKG energy)")
```

## 8. Visualisations

### 8.1 Energy timeline

```{r timeline-plot}
timeline_data <- rbind(synthetic_baseline, synthetic_workload)
timeline_data$cum_seconds <- cumsum(timeline_data$seconds)

plot_energy_timeline(timeline_data,
                     title = "Energy Consumption Over Time (synthetic data)")
```

### 8.2 Delta PKG violin plot

```{r delta-violin}
workload_with_deltas <- add_pop_dim_label(workload_with_deltas)

plot_delta_energy(workload_with_deltas,
                  geom      = "violin",
                  facet_col = "population_dimension",
                  title     = "Delta PKG Energy by Configuration (synthetic data)")
```

### 8.3 Delta PKG boxplot

```{r delta-boxplot}
plot_delta_energy(workload_with_deltas,
                  geom      = "boxplot",
                  facet_col = "population_dimension",
                  title     = "Delta PKG Energy by Configuration (synthetic data)")
```

## 9. LaTeX table formatting helpers

When preparing camera-ready papers, `format_mean_sd_latex()` produces the
familiar $\mu \pm \sigma$ notation in a form that can be passed directly to
`kableExtra::kable()` with `escape = FALSE`.

```{r latex-format}
wl_summary$pkg_latex <- format_mean_sd_latex(
  wl_summary$mean_delta_PKG,
  wl_summary$sd_delta_PKG
)
knitr::kable(wl_summary[, c("dimension", "population_size", "pkg_latex")],
             col.names = c("Dimension", "Population size", "Mean ± SD (J)"),
             caption   = "Delta PKG energy formatted for LaTeX output")
```

## 10. Working with real data

The vignette below outlines how the same workflow applies to the real
experimental data from the OLA-2026 paper.

```{r real-data, eval = FALSE}
# Download and load baseline data
baseline_file  <- download_bna("ola-base-ola-baseline-14-Dec-12-06-42.csv")
baseline_data  <- load_bna_csv(baseline_file)      # drops algorithm cols
baseline_summary <- summarize_baseline(baseline_data)

# Download and load workload data
workload_file   <- download_bna("ola-1.11.7-ola-14-Dec-13-02-30.csv")
workload_data   <- load_bna_csv(workload_file, drop_baseline_cols = FALSE)
workload_data   <- compute_deltas(baseline_summary, workload_data)

# Visualise
workload_data <- add_pop_dim_label(workload_data)
plot_delta_energy(workload_data, geom = "violin",
                  facet_col = "population_dimension")

# Statistical test
combined <- rbind(
  transform(baseline_data, work = "baseline"),
  workload_data
)
wilcoxon_tests(combined, "PKG")
```
